---
title: "Data transformation with dplyr"
subtitle: "Practice with penguins"
author: "Dominic Ziegler"
format: "html"
html:
embed-resources: true
editor: visual
execute:
  warning: false
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: load-packages
library(palmerpenguins)
library(dplyr)

```

All exercises in this assignment use the `penguins` data as a starting point.

1.  Run all code chunks above.

2.  Run the code chunk that contains `glimpse(penguins)`.

3.  How many variables are in the data set?

4.  How many observations are in the data set?

5.  What data types are contained in the variables? (Reminder: [https://rbtl-fs24.github.io/website/slides/lec-02-visualisation.html#/types-of-variables](https://ds4owd-001.github.io/website/slides/lec-02-visualisation.html#/types-of-variables){.uri})\
    \
    8 variables, 344 observations, \
    Variables:

    non numeric, categorical: species, island, sex\
    numeric, discrete: year

    numeric,continuous: the rest of them

```{r}
glimpse(penguins)
```

## Task 1: Create a subset of the data using filter()

Use `filter()` to create a subset from `penguins` that only contains observations for Adelie penguins.

```{r}
penguins_adelie <- filter(.data = penguins,species == "Adelie")
penguins_adelie
```

Use `filter()` to create a subset from `penguins` that only contains observations where body mass is less than or equal to 2900 g.

```{r}
penguins_m_small <- penguins %>% filter(body_mass_g <= 2900)
penguins_m_small
```

Use `filter()` to create a subset from `penguins` that only contains observations for Adelie penguins with a bill length greater than 40 mm.

```{r}
penguins_bl_large <- penguins %>% filter(bill_length_mm > 40)
penguins_bl_large
```

Use `filter()` to create a subset from `penguins` that excludes observations for chinstraps.

```{r}
penguins_nochin <- penguins %>% filter(species != "Chinstrap")
penguins_nochin
```

Use `filter()` to create a subset from `penguins` that only contains gentoo penguins with a bill depth greater than or equal to 15.5 millimeters.

```{r}
penguins_gentoo_bd <- penguins %>% filter(species == "Gentoo",bill_depth_mm >= 15.5)
penguins_gentoo_bd
```

Use `filter()` to create a subset from `penguins` that contains observations for male penguins recorded at Dream and Biscoe Islands.

```{r}
penguins_Dream_Biscoe_m <- penguins %>% filter(island == c("Dream","Biscoe"),sex == "male")
penguins_Dream_Biscoe_m
```

Use `filter()` to create a subset from `penguins` that contains observations for female Adelie penguins with bill lengths less than 35 mm.

```{r}
penguins_Adelie_f <- penguins %>% filter(species == "Adelie",
                                         bill_length_mm < 35,
                                         sex == "female")
penguins_Adelie_f
```

Use `filter()` to create a subset from `penguins` containing observations for female chinstrap penguins on Dream and Torgersen Islands.

```{r}
penguins_chin_D_T_f <- penguins %>% filter(sex == "female",
                                         species == "Chinstrap", 
                                         island == c("Dream","Torgersen"))
penguins_chin_D_T_f
```

Use `filter()` to create a subset from `penguins` that contains penguins that are either gentoos OR have a body mass greater than 4500 g.

```{r}
penguins_or <- penguins %>% filter(species == "Gentoo" | body_mass_g > 4500)
penguins_or
```

## Task 2: Add new columns with mutate()

Add a column to `penguins` that contains a new column `flipper_m`, which is the `flipper_length_mm` (flipper length in millimeters) converted to units of meters.

```{r}
penguins <- penguins %>% mutate(flipper_m = flipper_length_mm /1000)
glimpse(penguins)
```

Add a new column to `penguins` that contains a new column `body_mass_kg`, which is the `body_mass_g` (body mass in grams) converted to units of kilograms.

```{r}
penguins <- penguins %>% mutate(body_mass_kg = body_mass_g /1000)
```

Add a new column to `penguins` that contains a new column `bill_ratio`, which is the ratio of bill length to bill depth.

```{r}
penguins <- penguins %>% mutate(bill_ratio = bill_length_mm/bill_depth_mm)
```

Add a new column called id to `penguins` with a sequence of values from 1 to the length of the data frame. Use `relocate()` to move the column to the first position in the data frame.

```{r}
penguins <- penguins %>% 
  mutate(id = 1:n()) %>% 
  relocate(id,.before = everything())
```

## Task 3: Summarize data with group_by() and summarize() & count()

Starting with `penguins`, group the data by species, then create a summary table containing the maximum and minimum length of flippers (call the columns flip_max and flip_min). How will you handle NA values?

```{r}
penguins_sum <- penguins %>% 
  group_by(species) %>% 
  summarise(flip_max = max(flipper_length_mm, na.rm = TRUE),
            flip_min = min(flipper_length_mm, na.rm = TRUE))
penguins_sum
```

Starting with `penguins`, group the data by species and year, then create a summary table containing the mean bill depth (call this bill_depth_mean), the mean bill length (call this bill_length_mean), and the count for each group. How will you handle NA values?

```{r}
penguins_sum2 <- penguins %>% filter(!is.na(bill_length_mm) & !is.na(bill_depth_mm)) %>% 
  group_by(species,year) %>% 
  summarize(
    bill_depth_mean = mean(bill_depth_mm),
    bill_lenth_mean = mean(bill_length_mm),
    count = n()
  )
penguins_sum2
```

Use the `count()` function to count the number of observations for each species in `penguins`.

```{r}
penguins_sum3 <- penguins %>%
  group_by(species) %>%
  summarise(count = n())
penguins_sum3
```

Use the `count()` function to count the number of observations for each species and island in `penguins`.

```{r}
penguins_sum4 <- penguins %>%
  group_by(species,island) %>%
  summarise(count = n())
penguins_sum4
```

Use `filter()` to create a subset from `penguins` that contains observations for female penguins recorded at Torgersen and Biscoe Islands. Then use add the pipe `|>` and `count()` to verify that you written the correct code.

```{r}
penguins %>% filter(sex == "female",island == c("Torgersen","Biscoe")) %>% count(sex, island)

```

## Task 7: Data communication

**In the YAML header (between the three dashes at the top of the document)**

1.  Add your name as the author of this document
2.  Render the document and fix any errors

## Task 8: Stage, Commit & Push to GitHub

1.  Open the Git pane in RStudio. It's in the top right corner in a separate tab.
2.  **Stage** your changes by checking appropriate box next to all files (if you select one file with your mouse, you can then highlight them all with Ctrl + A on your keyboard and check all boxes).
3.  Write a meaningful commit message (e.g. "Completed part a of homework assignment 03.) in the **Commit message** box.
4.  Click **Commit**. Note that every commit needs to have a commit message associated with it.
